---
title: "question_5"
author: 'Chua Wei Yew Student Number: 190526948'
date: "3/31/2022"
output: html_document
---

Please download the files 2004.csv, 2005.csv, airports.csv, carriers.csv, and plane-data.csv from https://doi.org/10.7910/DVN/HG7NV7
Ensure that you have set your working directory to the location where the files are downloaded.

#Load libraries
```{r}
library(DBI)
library(dplyr)
```

#Create the Database
```{r}
setwd('F:/School/Year_3/ST2195/Course_Work/Data')
if (file.exists("airline2_r.db")){
  file.remove("airline2_r.db")
}

conn <- dbConnect(RSQLite::SQLite(), "airline2_r.db")
```

# ======== write to the database ========

# Load in the data from the csv files

```{r}
airports <- read.csv("airports.csv", header = TRUE)
carriers <- read.csv("carriers.csv", header = TRUE)
planes <- read.csv("plane-data.csv", header = TRUE)
dbWriteTable(conn, "airports", airports)
dbWriteTable(conn, "carriers", carriers)
dbWriteTable(conn, "planes", planes)

for(i in c(2004:2005)) {
  filename <- paste0(i, ".csv")
  print(paste("Processing:", i))
  ontime <- read.csv(filename, header = TRUE)
  if(i == 2004) {
    dbWriteTable(conn, "ontime", ontime)
  } else {
    dbWriteTable(conn, "ontime", ontime, append = TRUE)
  }
}
```

```{r}
ontime_db = tbl(conn, "ontime")
q5_df_ontime_db = data.frame(ontime_db)
q5_df_ontime_db = q5_df_ontime_db[q5_df_ontime_db$Cancelled != 1,] #removed flights that were cancelled, to lower number
#of entries and wont affect result
q5_df_ontime_db = q5_df_ontime_db[q5_df_ontime_db$Year != 2005,]
q5_df_ontime_db = q5_df_ontime_db[q5_df_ontime_db$Month == 1,]
```

```{r}
flight_sum_small = q5_df_ontime_db %>% group_by(Origin) %>% summarise(n_origin=n()) %>% arrange(desc(n_origin)) %>% head(10)
```

```{r}
flight_sum<- q5_df_ontime_db %>%
  #group flight cancellation and flight delay into one level
  mutate(Delay = ifelse(DepDelay >= 15 | is.na(DepDelay) == TRUE, 1, 0),
  ) %>%
  #select relevant variables and save to a new data table
  select(Delay, Year, Month, DayofMonth, DayOfWeek,Origin, CRSDepTime,
         CarrierDelay, WeatherDelay, NASDelay, SecurityDelay,LateAircraftDelay)
```

```{r}
flight_sum_small = q5_df_ontime_db %>% group_by(Origin) %>% summarise(n_origin=n()) %>% arrange(desc(n_origin)) %>% head(10)
```
We only select the top 10 Origins by using flight_sum_small.

```{r}
flight_sum = flight_sum[flight_sum$Origin %in% c('ORD','ATL','DFW','LAX','CVG','IAH','PHX','LAS','DEN','EWR'),]
flight_sum <- na.omit(flight_sum)

round(cor(flight_sum[,which(sapply(flight_sum,is.numeric))]), 3)
```
# SPLIT TRAINING AND TEST SETS DEP
```{r}
set.seed(100)
n <- nrow(flight_sum)
train_idx <- sample(1:n, ceiling(0.7*n))
length(train_idx)
```
```{r}
train <- flight_sum[train_idx, ]
test <- flight_sum[-train_idx, ]
dim(train)
```
# Use all predictors in the model
```{r}
mod.full <- glm(Delay ~ ., data = train, family = binomial)
summary(mod.full)
pchisq(mod.full$deviance,mod.full$df.residual,lower=FALSE)
```
# Random Forest
```{r}
library(randomForest)
mod.rf <- randomForest(Delay ~ ., data = train, ntree=100, prob=TRUE)
plot(mod.rf)
importance(mod.rf)
pred.rf <- predict(mod.rf, newdata = test, type = )
head(pred.rf)
```
```{r}
library(pROC)
roc_rf <- roc(test$Delay, pred.rf)
plot(roc_rf)
#Area under the ROC curve
roc_rf$auc
```